**Effective application of machine learning to microbiome based classification problems**

### 2019-10-14 submission/manuscript.Rmd
Jenna's edits last round

Running title: Machine learning methods in microbiome studies

@@

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training and evaluation methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a generalizable pipeline to train, validate and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and extreme gradient boosted trees (XGBoost). The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739]  but was slow to train (83.2 h) and difficult to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and was more interpretable. Our analysis showed that ML models should be chosen based on the goal of the study, as the choice will inform expectations of performance and interpretability.

@@

Prediction of health outcomes using machine learning (ML) is rapidly being adopted in microbiome studies. However, these ML models are likely overoptimistic in terms of quantifying predictive performance. Moreover, there is a trend towards using black box models such as random forest and neural networks without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease. This work represents a step towards developing better ML practices in microbiome research by implementing a rigorous pipeline and emphasizing the importance of selecting ML models that reflect the goal of the study. These concepts are not particular to the study of health outcomes but can also be applied to environmental microbiology studies.

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases. However, the structure of the human microbiome is remarkably variable among individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population seperately [@segata_metagenomic_2011]. Recently, machine learning models (ML) have grown in popularity among microbiome researchers because we are able to sample large numbers of individuals and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @mossotto_classification_2017]. The task of diagnosing an individual relies on a rigorously validated model. However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress. These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a modeling approach has been due to an implicit assumption that more complex models are better. This has resulted in a trend towards using models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although in some cases complex models may capture important non-linear relationships and therefore yield better predictions, they result in black boxes that lack interpretability. Such models require post hoc explanations to quantify the importance of each feature in making predictions. Depending on the goal of the modeling, researchers may choose to use different approaches. For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance. Nonetheless, it is important to understand the tradeoff and accept that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018]. It is important for researchers to justify their choice of modeling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing seven modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared their predictive performance, interpretability, and training time. To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classify individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset we developed a ML pipeline that can be used in many different scenarios for training and evaluating models. This framework can be easily applied to other host-associated and environmental microbiome datasets.

@@

First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset, while the test set was composed of the remaining 20% [Figure 1]. To maintain the distribution of controls and cases found in the full dataset, we performed stratified splits. For example, our full dataset included 490 individuals. Of these, 261 had normal colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%). A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals of which 52 had an SRN (54%). The training data was used to build and select the models and the test set was used for evaluating the model.

We trained seven different models using the training data [Table 1]. We focused on different classification algorithms and regularization methods. Regularization helps to prevent overfitting by penalizing a model that fits the training data too well. For regularized logistic regression and SVM with a linear kernel, we used L2-regularization to keep all potentially important features. For comparison, we also trained an L1-regularized SVM with a linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including a decision tree, a random forest, and XGBoost and an SVM with a non-linear kernel.

Model selection required tuning hyperparameters. Hyperparameters are parameters which need to be specified or tuned by the user, in order to train a model for a specific modeling problem. For example, when using regularization, C is a hyperparameter that indicates the penalty for overfitting. Hyperparameters are tuned using the training data to find the best model. We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2]. The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls. An AUROC value of 0.5 indicates that model's predictions are no different than random. To select hyperparameters, we performed a full grid search for hyperparameter settings when training the models. Default hyperparameter settings in previously developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task. For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and needed to be tuned to find the optimal C value [Figure S1].

Once hyperparameters were selected, we trained the model using the full training dataset and applied the final model to the held-out data to evaluate the testing predictive performance of each model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to obtain a reliable and robust interpretation of model performance [Figure 1].

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had higher test AUROC values than the other models for detecting SRNs (resampling test between means of two groups, p < 0.01). The median AUROC of the random forest model was `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`]. L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models. To support our claim, we investigated the distribution of differences between random forest and L2-regularized logistic regression AUROC values for each data-split (Figure S3). Our results underscored that L2-regularized logistic regression perform nearly as well as random forest.

@@

__Interpretation of each ML model.__ Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019]. Because we often use ML models not just to predict a health outcome but also to identify the biomarkers for a disease, model interpretation becomes crucial for microbiome studies. ML models decrease in interpretability as they increase in complexity. In this study we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3]. We also reviewed the signs of feature weights to determine whether an OTU was associated with classifying a subject as being healthy or having an SRN. It was encouraging that many of the highest ranked OTUs were shared across these three models, (e.g. OTUs 50, 426, 609, 822, 1239). The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU. However, this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.  In this case, similar to previous studies [@sze_leveraging_2018; @wirbel_meta-analysis_2019], it was not possible to distinguish between health and disease using a single OTU. Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU was associated with classifying a subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S5]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S5], 17 of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. Although permutation analysis does not not allow us to determine the weight or the sign of the features, these results indicate that both methods are consistent in selecting the most important OTUs.

__The computational efficiency of each ML model.__ We compared the training times of the seven ML models. As expected, the training times increased with the complexity of the model and the number of potential hyperparameter combinations. Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].

@@

To address these needs, we developed an open-sourced framework to ML models. Using this pipeline, we benchmarked seven ML models and showed that the tradeoff between model complexity and performance may be less severe than originally hypothesized. In terms of predictive performance, the random forest model had the best AUROC compared to the other six models. However, the second-best model was L2-regularized logistic regression with a median AUROC difference less than 0.015 compared to random forest. While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest is a non-linear ML model and is most often explained using post hoc methods such as permutation importance. On the other hand, using L2-regularized logistic regression we ranked the importance of each OTU based on their feature weights. Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.

We established a pipeline that can be generalized to any modeling method that predicts a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this data-split 100 times to measure the possible variation in predictive performance. During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the good test AUROCs.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other, however, we could reduce the correlation threshold to further investigate the relationships among correlated features. It is important to know the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.

@@

Our framework provides a reproducible structure to investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant. However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly harder and distinct undertaking. For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data. We are also challenged by the practical constraints of OTU-based approaches. The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added [@westcott_opticlust_2017]. Finally, we also need independent validation cohorts to test the performance of a diagnostic model. To realize the potential for using ML approaches with microbiome data, it is necessary that we direct our efforts to overcome these challenges.

>>>>

Running title: Machine learning methods applied to microbiome studies

@@

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu, wiensj@umich.edu

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training and evaluation methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a reusable open-source pipeline to train, validate and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and gradient boosted trees (XGBoost). The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739]  but was slow to train (83.2 h) and not immidiately interpretable. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and yielded a model that was inherently interpretable. Our analysis highlights the importance of choosing an ML approach should be chosen based on the goal of the study, as the choice will inform expectations of performance and interpretability.

@@

Prediction of health outcomes using machine learning (ML) is rapidly being adopted in microbiome studies. However, estimated performance associated with these ML models is likely overoptimistic. Moreover, there is a trend towards using black box models without a discussion of the difficulty of interpreting such models when trying to identify microbial biomarkers of disease. This work represents a step towards developing more reproducble ML practices in applying ML to microbiome research. We implement a rigorous pipeline and emphasize the importance of selecting ML models that reflect the goal of the study. These concepts are not particular to the study of health outcomes but can also be applied to environmental microbiology studies.

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose disease. However, the structure of the human microbiome is remarkably variable among individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population seperately [@segata_metagenomic_2011]. Recently, machine learning (ML) models have grown in popularity among microbiome researchers as our ability to to sample large numbers of individuals has grown; such models can effectively account for the interpersonal microbiome variation and the ecology of disease.

ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @mossotto_classification_2017]. The task of diagnosing an individual relies on a rigorously validated model. However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress. These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances. Though, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to further improve reproducibility and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a modeling approach has often been due to an implicit assumption that more complex models are better. This has resulted in a trend towards using non-linear models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although in some cases complex models may capture important non-linear relationships and therefore yield better predictions, they can also result in black boxes that lack interpretability. Such models require post hoc explanations to quantify the importance of each feature in making predictions. Depending on the goal of the modeling, other approaches may be more appropriate. For example, researchers trying to identify the microbiota associated with disease may desire a more interpretable model, whereas clinicians may (but not always) emphasize predictive performance. Nonetheless, it is important to understand that the benefit of more complex, less interpretable models may be minimal [@rudin_please_2018; @rudin_optimized_2018]. It is important for researchers to justify their choice of modeling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing seven modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and gradient boosted trees. We compared their predictive performance, interpretability, and training time. To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classify individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset, we developed an ML pipeline that can be used in many different scenarios for training and evaluating models. This framework can be easily applied to other host-associated and environmental microbiome datasets.

@@

First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset, while the test set was composed of the remaining 20% [Figure 1]. To maintain the distribution of controls and cases found in the full dataset, we performed stratified splits. For example, our full dataset included 490 individuals. Of these, 261 had normal colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%). A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals of which 52 had an SRN (54%). The training data were used to build and select the models and the test set was used for evaluating the model.

We trained seven different models using the training data [Table 1]. We focused on different classification algorithms and regularization methods. Regularization helps to prevent overfitting by penalizing a model that fits the training data too well [@Ng:2004:FSL:1015330.1015435]. For regularized logistic regression and SVM with a linear kernel, we used L2-regularization to keep all potentially important features. For comparison, we also trained an L1-regularized SVM with a linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including a decision tree, a random forest, and gradient boosted trees (XGBoost) and an SVM with a non-linear kernel.

Model selection requires tuning hyperparameters. Hyperparameters are parameters that need to be specified or tuned by the user, in order to train a model for a specific modeling problem. For example, when using regularization, C is a hyperparameter that indicates the penalty for overfitting. Hyperparameters are tuned using the training data to find the best model. We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2]. The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls. An AUROC value of 0.5 indicates that model's predictions are no different than random. To select hyperparameters, we performed a grid search for hyperparameter settings when training the models. Default hyperparameter settings in previously developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and should to be optimized for each new ML task. For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning [Figure S1].

Once hyperparameters were selected, we trained the model using the full training dataset and applied the final model to the held-out data to evaluate the testing predictive performance of each model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to obtain a robust interpretation of model performance, less likely to be affected by a "lucky" or "unlucky" split [Figure 1].

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had a median AUROC value of `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`] which was higher than the other models for detecting SRNs. However, we determined that the random forest model performance was not statistically significantly higher (p = 0.5) when we investigated the distribution of differences between random forest and L2-regularized logistic regression AUROC values for each data-split (Figure S3). Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well non-linear models.

@@

__Interpretation of each ML model.__ Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019]. Because we often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease, model interpretation becomes crucial for microbiome studies. ML models often decrease in interpretability as they increase in complexity. In this study, we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3]. We also reviewed the signs of feature weights to determine whether an OTU was associated with classifying a subject as being healthy or having an SRN. It was encouraging that many of the highest ranked OTUs were shared across these three models, (e.g. OTUs 50, 426, 609, 822, 1239). The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU. However, this approach is limited to linear models or models with prespecified interaction terms.

Second, to analyze non-linear models we interpreted the feature importance using permutation importance [@10.1093/bioinformatics/btq134]. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.  Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post hoc (i.e., done using the test data).

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S5]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S5], 17 of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. These results indicate that both methods are consistent in selecting the most important OTUs.

__The computational efficiency of each ML model.__ We compared the training times of the seven ML models. We did not take advantage of parallelization in training nor was the code optimized for speed. Still, as expected, the training times increased with the complexity of the model and the number of potential hyperparameter combinations. Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].

@@

To address these needs, we developed an open-sourced framework to ML models. Using this pipeline, we benchmarked seven ML models and showed that the tradeoff between model complexity and performance may be less severe than originally hypothesized. In terms of predictive performance, the random forest model had the best AUROC compared to the other six models. However, the second-best model was L2-regularized logistic regression with a median AUROC difference less than 0.015 compared to random forest. While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest is a non-linear ML model while L2-regularized logistic regression, a linear model, is easily interpreted according to the feature weights. Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.

We established a pipeline that can be generalized to any modeling method that predicts a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this data-split 100 times to measure the possible variation in predictive performance. During training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline generalized to the held-out test sets.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we partially accounted for collinearity by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other, however, we could reduce the correlation threshold to further investigate the relationships among correlated features. It is important to know the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. To identify the true underlying microbial factors of a disease, it is crucial to follow up any correlation analyses with further hypothesis testing and experimentation for biological validation.

@@

Our framework provides a reproducible structure to investigators wanting to train, evaluate, and interpret their own ML models to generate hypotheses regarding which OTUs might be biologically relevant. However, deploying microbiome-based models to make clinical diagnoses or predictions is a significantly more challenging and distinct undertaking. For example, we currently lack standardized methods to collect patient samples, generate sequence data, and report clinical data. We are also challenged by the practical constraints of OTU-based approaches. The de novo algorithms commonly in use are slow, require considerable memory, and result in different OTU assignments as new data are added [@westcott_opticlust_2017]. Finally, we also need independent validation cohorts to test the performance of a diagnostic model. To realize the potential for using ML approaches with microbiome data, it is necessary that we direct our efforts to overcome these challenges.
